---
title: "Math_564__FINAL_PROJECT"
output:
  pdf_document: default
  html_document: default
  fig_width: 200
  fig_height: 200
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Read the data and create the transaction month column and convert variables to factors 
```{r}
library(readxl)
d <- read_excel("C:\\Users\\sagar\\OneDrive\\Documents\\STAT425FP\\Real_estate_valuation_data_set.xlsx")
d$`X7 transaction month` = (round((d$`X1 transaction date`-floor(d$`X1 transaction date`))*12, digits = 0)+1)
d$`X7 transaction month`= as.factor(d$`X7 transaction month`)
d
```
##Column names of data for section 2 
```{r}
names(d)
```

##Structure of the data for section 2 
```{r}
str(d)
```


##Summary Statistics on the different variables for Section 2

```{r}
summary(d)
```
```{r}
pairs(~`X1 transaction date`+`X2 house age`+`X3 distance to the nearest MRT station`+`X4 number of convenience stores`+`X5 latitude`+`X6 longitude`+`X7 transaction month`+`Y house price of unit area`, data = d)
```


##Plotting different variables against the house prices for Section 2 
```{r}
par(mfrow=c(3,3))
plot(d$`X1 transaction date`, d$`Y house price of unit area`)
plot(d$`X2 house age`,d$`Y house price of unit area`)
plot(d$`X3 distance to the nearest MRT station`,d$`Y house price of unit area`)
plot(d$`X4 number of convenience stores`,d$`Y house price of unit area`)
plot(d$`X5 latitude`, d$`Y house price of unit area`)
plot(d$`X6 longitude`,d$`Y house price of unit area`)
plot(d$`X7 transaction month`,d$`Y house price of unit area`)

```
```{r}
library(ggplot2)
library(GGally)

```

```{r}
ggpairs(d, columns=c(2:9)) +
ggtitle("Data Set Plot 2")
```


#train test split of the data 
```{r}
train= d[1:290,]  
test= d[291:414,]
```

#MLR using backwards selection in Section 3 
```{r}
m = lm(`Y house price of unit area`~`X1 transaction date`+`X2 house age`+`X3 distance to the nearest MRT station`+`X4 number of convenience stores`+`X5 latitude`+`X6 longitude`+`X7 transaction month`, data = train)
```

```{r}
summary(m)
```

```{r}
bm = step(m,direction="backward")
```

```{r}
summary(bm)
```

##Checking for high leverage points 
```{r}
library(faraway)
n = 291; p = 5;
lev = influence(bm)$hat
lev[lev>2*p/n]
```
```{r}
halfnorm(lev, 6, labs = row.names(train), ylab = "lev")
```
```{r}
jack=rstudent(bm); 
qt(.05/(2*n), 285)
sort(abs(jack), decreasing=TRUE)[1:6]
```
```{r}
cook = cooks.distance(bm)
max(cook)
halfnorm(cook, labs=row.names(train), ylab="Cook's distances")
```

```{r}
t1 = train[row.names(train) != 271,]
m2 = lm(`Y house price of unit area`~`X1 transaction date`+`X2 house age`+`X3 distance to the nearest MRT station`+`X4 number of convenience stores`+`X5 latitude`+`X6 longitude`+`X7 transaction month`, data= t1)
bm2 = step(m2,direction="backward")
summary(bm2)
```

## Running test for homoedascity

```{r}
library(lmtest)
bptest(bm2)
```
# Running test for normality 

```{r}
shapiro.test(residuals(bm2))
```
#Running tests for linearity 

```{r}
r.y2 = update(bm2, ~. - `X2 house age`)$res;
r.x2 = lm(`X2 house age` ~., data=t1[,-1])$res;
tmp1=lm(r.y2 ~ r.x2); summary(tmp1)
```

```{r}
r.y3 = update(bm2, ~. - `X3 distance to the nearest MRT station`)$res;
r.x3 = lm(`X3 distance to the nearest MRT station` ~., data=t1[,-1])$res;
tmp2=lm(r.y3 ~ r.x3); summary(tmp2)
```

```{r}
r.y4 = update(bm2, ~. - `X4 number of convenience stores`)$res;
r.x4 = lm(`X4 number of convenience stores` ~., data=t1[,-1])$res;
tmp3=lm(r.y4 ~ r.x4); summary(tmp3)
```

```{r}
r.y4 = update(bm2, ~. - `X5 latitude`)$res;
r.x4 = lm(`X5 latitude` ~., data=t1[,-1])$res;
tmp4=lm(r.y4 ~ r.x4); summary(tmp4)
```

#Calculating testing error
```{r}
library(Metrics)
test1 = subset(test, select = -`Y house price of unit area`)
p = predict(bm2, test1)
data = data.frame(pred = p, actual = test$`Y house price of unit area`)
mean((data$actual - data$pred)^2)
rmse(data$pred, data$actual)
```

##Section 3.3 MARS 
```{r}
library(earth)
m3 = earth(`Y house price of unit area`~`X2 house age`+`X3 distance to the nearest MRT station`+`X4 number of convenience stores`+`X5 latitude`, data= t1)
summary(m3)
```

##Prediction Error for MARS
```{r}
p1 = predict(m3, test1)
rmse(p1, test$`Y house price of unit area`)
```

